\chapter{Communalism: The Liberal Phase}



There was hardly any communalism in India before the last quarter of the 19th century. As is well-known, Hindus and Muslims had fought shoulder to shoulder in the Revolt of 1857. The notion of Hindu-Muslim distinction at the non-religious plane, not to speak of the clash of interests of Hindus and Muslims was by and large non-existent in the Press during the 1860s. The identity that the North Indian newspapers emphasised was that of the Hindustanis, especially vis-a-vis European or British rulers 

Even when some Muslim intellectuals began to notice that Muslims in some parts of the country were tagging behind Hindus in modern education and in government jobs, they blamed not Hindus but the Government's anti-Muslim policy and the neglect of modem education by upper class Muslims. Syed Ahmed Khan, undoubtedly one of the outstanding Indians of the l9thiitury, began his educational activities without any communal bias. The numerous scientific societies he founded in 1860s involved both Hindus and Muslims. The Aligarh College he specially founded to fight the bias against modern education among Muslims, received financial support from moneyed Hindus; and its faculty and students had a large Hindu component Syed Ahmed loudly preached the commonness of Hindus and Muslims till the founding of the Congress in 1885. Thus, for example, he said in 1884: `Do you not inhabit this land? Are you not buried in it or cremated on it? Surely you live and die on the same land. Remember that Hindus and Muslims are religious terms. Otherwise Hindus, Muslims and Christians who live in this country are by virtue of this fact one qawm' (nation or community). 

Ironically, communalism in India got its initial start in the 1 880s when Syed Ahmed Khan counterposed it to the national movement initiated by the National Congress. In 1887, Dufferin, the Viceroy, and A. Colvin the Lieutenant-Governor of U.P., launched a frontal public attack on the National Congress, once its anti-imperialist edge became clear. Syed Ahmed, believing that the Muslims' share in administrative posts and in profession could be increased only by professing and proving loyalty to the colonial rulers, decided to join in the attack. Furthermore, he felt that he needed the active support of big zamindars and the British officials for the Aligarh College. Initially he made an attempt with the help of Shiva Prasad, Raja of Bhinga, and others to organize along caste, birth, class and status lines the feudal (jagirdari) and bureaucratic elements in opposition to the rising democratic national movement. However, this attempt failed to get off the ground. 

Syed Ahmed now set out to organize the jagirdari elements among Muslims as Muslims or the Muslim qawm (community). He and his fo1lowrs gradually laid down the foundation of all the basic themes of the communal ideology as it was to be propagated in the first half of the 20th century. A basic theme was that Hindus, because they were a majority, would dominate Muslims and `totally override the interests of the smaller community' if representative, democratic government was introduced or if British rule ended and power was transferred to Indians. The British were needed to safeguard Muslims as a minority. In the Indian context, said Syed Ahmed, they were the best guardians of Muslim interests. Muslims must, therefore, remain loyal and oppose the National Congress. The theme of a permanent clash of `Hindu' and `Muslim' interests was also brought forth. Giving up his earlier views, he now said that India could not be considered a nation. He declared that the Congress was a Hindu body whose major objectives were `against Muslim interest.' Simultaneously, he criticized the Congress for basing itself on the principle of social equality among the `lowly' and the `highly' born. Objecting to the Congress demand for democratic elections, Syed Ahmed said that this would `mean that Muslims would not be able to guard their interests, for 9t would be like a game of dice in which one man had four dice and the other only one.' Any system of elections, he said, would put power into the hands of `Bengalis or of Hindus of the Bengali type' which would lead to Muslims falling into `a condition of utmost degradation' and `the ring of slavery' being put on them by Hindus. Syed Ahmed and his co-workers also demanded safeguards for Muslims in Government jobs, legislative councils, and district boards and recognition of the historical role and political importance of Muslims so that their role in legislative councils should not be less than that of Hindus. At the same time, Syed Ahmed and his followers did not create a counter command political organization, because the British authorities at the time frowned upon any politicization of the Indian people. Syed Ahmed held that any agitational politics would tend to become anti- government and seditious and to create suspicion disloyalty among the rulers. He, therefore, asked Muslims to shun all polities and remain politically passive, i.e., non-agitational, in their approach. The co1onial rulers were quick to see the inherent logic or communalism and the theory of the official protection of the minorities and from the beginning actively promoted and supported communalism. 

The Muslim communalists continued to follow the politics of loyalty after Syed Ahmed's death. They openly sided with the Government during the Swadeshi Movement in Bengal during 1905-6 and condemned the Muslim supporters of the movement as `vile traitors' to Islam and as `Congress touts.' But the attempt to keep the growing Muslim intelligentsia politically passie or loyalist was not wholly successful. Badruddin Tyabji presided over the Congress session in 1887, and the number of Muslim delegates to the Congress increased in the succeeding years. R.M. Sayani, A. Bhimji, Mir Musharaff Hussain, Hamid Ali Khan and numerous other Muslim intellectuals from Bombay, Bengal and Northern India joined the Congress. They pointed out that not even one of the Congress demands was communal or for Hindus only. The nationalist trend continued to spread among Muslims all over the country till the end of the 19th century. Abdul Rasul and a large number of other Bengali Muslim intellectuals gave active support to the Swadeshi agitation against the partition of Bengal. In fact, the nationalist trend remained dominant among Muslims in Bengal till the late 1920s. 

Once the Swadeshi Movement brought mass politics to India, a large section of the Muslim intelligentsia could not be kept away from the Congress; the British Government felt compelled to difer some constitutional concessions, and it became impossible to continue to follow the policy of political passivity. The communalists, as also their official supporters, felt that they had to enter the political arena. At the end of 1907 the 

All India Muslim League was founded by a group of big zamindars ex-bureaucrats and other upper class Muslims like the Aga Khan, the Nawab of Dacca and Nawab Mohsin-uI-Muk. Founded as a loyalist, communal and conservative political organization, the League supported the partition of Bengal, raised the slogan of separate Muslim interests, demanded separate electorates and safeguards for Muslims in government services, and reiterated all the major themes of communal politics and ideology enunciated earlier by Syed Ahmed and his followers. Viqar-ul-Mulk for example, said: `God forbid, if the British rule disappears from India, Hindus will lord over it; and we will be in constant danger of our life, property and honour. The only way for the Muslims to escape this danger is to help in the continuance of the British rule.'9 He also expressed the fear `of the minority losing its identity.' One of the major objectives of the Muslim League was to keep the emerging intelligentsia among Muslims from joining the Congress. Its activities were directed against the National Congress and Hindus and not against the colonial regime. 

Simultaneously, Hindu communalism, was also being born. From the 1870s, a section of Hindu zamindars, moneylenders and middle class professionals began to arouse anti-Muslim sentiments. Fully accepting the colonial view of Indian history, they talked of the `tyrannical' Muslim rule in the medieval period and the `liberating' role of the British in `saving' Hindus from `Muslim oppression.' In U.P. and Bihar, they took up the question of Hindi and gave it a communal twist, declaring that Urdu was the language of Muslims and Hindi of Hindus. All over India, anti-cow slaughter propaganda was undertaken in the early 1890s, the campaign being primarily directed not against the British but against Muslims; the British cantonments, for example, were left free to carry on cow slaughter on a large scale. Consequently, this agitation invariably took a communal turn, often resulting in communal riots. The anti-cow slaughter agitation died down by 1896, to be revived again in a more virulent form in the second decade of the 20th century. The Hindu communalists also carried on a regular agitation for a `Hindu' share of seats in legislatures and in government services. 

The Punjab Hindu Sabha was founded in 1909. Its leaders, 

U.N. Mukherji and Lal Chand, were to lay down the foundations of Hindu communal ideology and politics. They directed their anger primarily against the National Congress for trying to unite Indians into a single nation and for `sacrificing Hindu interests' to appease Muslims. In his booklet, Self-Abnegation in Politics, Lal Chand described the Congress as the `self-inflicted misfortune' of Hindus. Hindus, he wrote, were moving towards extinction because of `the poison imbibed for the last 25 years.' They could be saved only if they were willing to `purge' the poison and get rid of the `evil.' He accused the Congress of making `impossible' demands on the Government, leading to its justifiable anger against the Congress and Hindus. Instead Hindus should try to neutralize the third party, the Government, in their fight against Muslims. It was also essential that Hindus abandon and `end' the Congress. `A Hindu,' Lal Chand declared, `should not only believe but make it a part and parcel of his organism, of his life and of his conduct, that he is a Hindu first and an Indian after.'' 

The first session of the All-India Hindu Mahasabha was held in April 1915 under the presidentship of the Maharaja of Kasim Bazar. But it remained for many years a rather sickly child compared to the Muslim League. This was for several reasons. The broader social reason was the greater and even dominant role of the zamindars, aristocrats and ex-bureaucrats among Muslims in general and even among the Muslim middle classes. While among Parsis and Hindus, increasingly, it was the modern intelligentsia, with its emphasis on science, democracy and nationalism, and the bourgeois elements in general, which rapidly acquired intellectual, social, economic and political influence and hegemony, among Muslims the reactionary landlords and mullahs continued to exercise dominant influence or hegemony. Landlords and traditional religious priests, whether Hindu or Muslim, were conservative and supporters of established, colonial authority. But while among Hindus, they were gradually losing positions of leadership, they continued to dominate among Muslims. In this sense the weak position of the middle class among Muslims and its social and ideological backwardness contributed to the growth of Muslim communalism. 

There were other reasons for the relative weakness of Hindu communalism. The colonial Government gave Hindu communalism few concessions and little support, for it banked heavily on Muslim communalism and could not easily simultaneously placate both communalisms. 

The colonial authorities and the communalists together evolved another powerful instrument for the spread and consolidation of communalism in separate electorates which were introduced in the Morley-M into Reforms of 1907. Under this system, Muslim voters (and later Sikhs and others) were put in separate constituencies from which only Muslims could stand as candidates and for which only Muslims could vote. Separate electorates turned elections and legislative councils into arenas for communal conflicts. Since the voters were exclusively the followers of one religion, the candidates did not have to appeal to voters belonging to other religions. They could, therefore, make blatantly communal appeals and voters and others who listened to these appeals were gradually trained to think and vote communally and in general to think in terms of `communal' power and progress and to express their socio-economic grievances in communal terms. The system of reservation of seats and weightage in legislatures, government services, educational institutions etc., also had the same consequences. 

A slight detour at this stage is perhaps necessary. When discussing the history of the origins and growth of communalism and communal organizations, one particular error is to be avoided. Often a communalist ascribed — or even now ascribes in historical writings — the origins of one communalism to the existence of and as a reaction to the other communalism. Thus, by assigning the `original' blame to the other communalism a sort of backdoor justification for one's own communalism is (or was) provided. Thus the Hindu, Muslim or Sikh communalists justified their own communalism by arguing that they were reacting to the communalism initiated by others. In fact, to decide which communalism came first is like answering the question: which came first, the chicken or the egg? Once communalism arose and developed, its different variants fed and fattened on each other. 

The younger Muslim intellectuals were soon dissatisfied with the loyalist, anti-Hindu and slavish mentality of the upper class leadership of the Muslim League. They were increasingly drawn to modern and radical nationalist ideas. The militantly nationalist Ahrar movement was founded a this time under the leadership of Maulana Mohammed Au, Hakim Ajmal Khan, Hasan Imam, Maulana Zafar Ali Khan, and Mazhar-uI-Haq. In their efforts, they got support from a section of orthodox uiwna scholars) especially those belonging to the Deoband school. Another orthodox scholar to be attracted to the national movement was the young Maulana Abul Kalam Azad, who was educated at the famous Al Azhar University at Cairo and who propagated his rationalist and nationalist ideas in his newspaper Al Hilal which he brought out in 1912 at the age of twenty-four After an intense struggle, the nationalist young Muslims came to the fore in the Muslim League. They also became active in the Congress. In 1912, the brilliant Congress leader, M.A. Jinnah, was invited to join the League which adopted self-government as one of its objectives, in the same year, the Aga Khan resigned as the President of the League. 

From 1912 to 1924, the young nationalists began to overshadow the loyalists in the League which began to move nearer to the policies of the Congress. Unfortunately, their nationalism was flawed in so far as it was not fully secular (except with rare exceptions like Jinnah). It had a strong religious and pan-Islamic tinge. Instead of understanding and opposing the economic and political consequences of modern imperialism they fought it on the ground that it threatened the Caliph (khalifa) and the holy places. Quite often their appeal was to religious sentiments. This religious tinge or approach did not immediately clash with nationalism. Rather, it made as adherents anti-imperialist; and it encouraged the nationalist trend among urban Muslims. But in the long run it proved harmful as it inculcated arid encouraged the habit of looking at political questions from a religious point of view. The positive development within the Congress — discussed in an earlier CHAPTER - and within the Muslim League soon led to broad political unity among the two, an important role in this being played by Lokmanya Tilak and M.A. Jinnah. The two organizations held their sessions at the d of 1916 at Lucknow, signed a pact known as the Lucknow Pact, and put forward common political demands before the Government including the demand for self-government for India after the war. The Pact accepted separate electorates and the system of weightage and reservation of seats for the minorities in the legislatures. While a step forward in many respects — and it enthused the political Indian — the Pact was also a step back. The Congress had accepted separate electorates and formally recognized communal politics. Above all, the Pact was tacitly based on the assumption that India consisted of different communities with separate interests of their own. It, therefore, left the way open to the future resurgence of communalism in Indian politics. 

The nationalist movement and Hindu-Muslim unity took giant steps forward after World War I during the agitation against the Rowlatt Acts, and the Khilafat and the Non-Cooperation Movements. As if to declare before the world the principle of Hindu-Muslim unity in political action, Swami Shradhanand, a staunch Arya Samajist, was asked by Muslims to preach from the pulpit of the Jama Masjid at Delhi, while Dr. Saifuddin Kitchlu, a Muslim, was given the keys to the Golden Temple, the Sikh shrine at Amritsar. The entire country resounded to the cry of `Hindu-Muslim ki Jai'. The landlord-communalists and ex- bureaucrats increasingly disassociated themselves from the Muslim League, while the League itself was overshadowed by the Khilafat Committee as many of the League leaders — as also many of the old Congress leaders — found it difficult to keep pace with the politics of a mass movement. Even though the Khilafat was a religious issue, it resulted in raising the national, anti- imperialist consciousness of the Muslim masses and middle classes. Moreover, there was nothing wrong in the nationalist movement taking up a demand that affected Muslims only, just as the Akali Movement affected the Sikhs only and the anti­ untouchability campaign Hindus only. 

But there were also certain weaknesses involved. The nationalist leadership failed to some extent in raising the religious political consciousness of Muslims to the higher plane of secular political consciousness. The Khilafat leaders, for example, made appeals to religion and made full use of fatwas (opinion or decision on a point of Islamic law given by a religious person of standing) and other religious sanctions. Consequently, they strengthened the hold of orthodoxy and priesthood over the minds of men and women and encouraged the habit of looking at political questions from the religious point of view. By doing so and by emphasizing the notion of Muslim solidarity, they kept an opening for communal ideology and politics to grow at a later stage. 

The Non-Cooperation Movement was withdrawn in February 1922. As the people felt disillusioned and frustrated and the Dyarchy became operational, communalism reared its ugly head and in the post-1922 years the country was repeatedly plunged into communal riots. Old communal organizations were revived and fresh ones founded. The Muslim League once again became active and was cleansed of radical and nationalist elements. The upper class leaders with their open loyalism and frankly communal ideology once again came to the fore. The Hindu Mahasabha was revived in 1923 and openly began to cater to anti-Muslim sentiments. Its proclaimed objective became `the maintenance, protection and promotion of Hindu race, Hindu culture and Hindu civilization for the advancement of Hindu Rashtra.' The Hindu as well as Muslims communalists tried to inculcate the psychology of fear among Hindus and Muslims — the fear of being deprived, surpassed, threatened, dominated, suppressed, beaten down, and exterminated. It was during these years that Sangathan and Shuddhi movements among Hindus and Tanzeem and Tabligh movements among Muslims, working for communal consolidation and religious conversion, came up. The nationalists were openly reviled as apostates and as enemies of their own religion and co-religionists. A large number of nationalists were not able to withstand communal pressure and began to adopt communal or semi- communal positions. The Swarajists were split by communalism. A group known as `responsivists' offered cooperation to the Government so that the so-called Hindu interests might be safeguarded. Lajpat Rai, Madan Mohan Malaviya and N.C. Kelkar joined the Hindu Mahasabha and argued for Hindu communal solidarity. The less responsible `responsivists' and Hindu Mahasabhaites carried on a virulent campaign against secular Congressmen. They accused MOWSI Nehru of letting down Hindus, of being anti-Hindu and an Islam- lover, of favo.xmg cow-slaughter, and of eating beef. Many old Khilafatists also now turned communal. The most dramatic shift was that of Maulanas 

Mohammed All and Shaukat All who now accused the Congress of trying to establish a Hindu Government and Hindus of wanting to dominate and suppress Muslims. The most vicious expression of communalism were communal riots which broke out in major North Indian cities during 1923-24. According to the Simon Commission Report, nearly 112 major communal riots occurred between 1922 and 1927. 

The nationalist 1eadership made strenuous efforts to oppose communal political forces, but was not able to evolve an effective line of action. What was the line of action that it adopted and why did it fail? Its basic strategy was to try to bring about unity at the top with communal leaders through negotiations. This meant that either the Congress leaders acted as mediators or intermediaries between different communal groups or they themselves tried to arrive at a compromise with Muslim communal leaders on questions of `protection' to and `safeguards' of the interests of the minorities in terms of reservation of seats in the legislatures and of jobs in the government. 

The most well-known of such efforts was made during 1928. As an answer to the challenge of the Simon Commission, Indian political leaders organized several all-India conferences to settle communal issues and draw up an agreed constitution for India. A large number of Muslim communal leaders met at Delhi in December 1927 and evolved four basic demands known as the Delhi Proposals. These proposals were: (1) Sind should be made a separate province; (2) the North-West Frontier Province should be treated constitutionally on the same footing as other provinces; Muslims should have 33 1/3 per cent representation in the central legislature; (4) in Punjab and Bengal, the proportion of representation should be in accordance with the population, thus guaranteeing a Muslim majority, and in other provinces, where Muslims were a minority, the existing reservation of seats for Muslims should continue. 

The Congress proposals came in the form of the Nehru Report drafted by an all-parties committee. The Report was put up for approval before an All-Party Convention at Calcutta at the end of December 1928. Apart from other aspects, the Nehru Report recommended that India should be a federation on the basis of linguistic provinces and provincial autonomy, that elections be held on the basis of joint electorates and that seats in central and provincial legislatures be reserved for religious minorities in proportion to their population. The Report recommended the separation of Sind from Bombay and constitutional reform in the North-West Frontier Province. 

The Report could not be approved unanimously at the Calcutta Convention. While there were wide differences among Muslims communalists. a section of the League and the Khilafatists were willing to accept joint electorates and other proposals in the Report provided three amendments, moved by 

M.A. Jinnah, were accepted. Two of these were the same as the third and fourth demands in the Delhi Proposals, the first and the second of these demands having been conceded by the Nehru Report. The third was a fresh demand that residuary powers should vest in the provinces. A large section of the League led by Mohammed Shafi and the Aga Khan and many other Muslim communal groups refused to agree to these amendments; they were not willing to give up separate electorates. The Hindu Mahasabha and the Sikh League raised vehement objections to the parts of the Report dealing with Sind, North-West Frontier Province, Bengal and Punjab. They also refused to accept the Jinnah amendments. The Congress leaders were not willing to accept the weak centre that the Jinnah proposals envisioned. 

Most of the Muslim communalists now joined hands and Jinnah too decided to fall in line. Declaring that the Nehru Report represented Hindu interests, he consolidated all the communal demands made by different communal organizations at different times into a single document which came to be known as Jinnah's Fourteen Points. The Fourteen Points basically consisted of the four Delhi Proposals, the three Calcutta amendments and demands for the continuation of separate electorates arid reservation of seats for Muslims in government services and self- governing bodies. The Fourteen Points were to form the basis of all future communal propaganda in the subsequent years. 

This strategy of trying to solve the communal problem through an agreement or pact with the Hindu, Muslim and Sikh communal leaders proved a complete failure and suffered from certain inherent weaknesses. Above all it meant that the 

Congress tacitly or by implication accepted, to a certain extent, the claim of the communal leaders that they were representatives of the communal interests of their respective `communities,' and, of course, that such communal interests and religious communities existed in real life. By negotiating with communal leaders, the Congress legitimized their politics and made them respectable. It also weakened its right, as well as the will, to carry on a hard political-ideological campaign against communal parties and individuals. Constant negotiations with Muslim communal leaders wakened the position of secular, anti- imperialist Muslims and Muslim leaders like Azad, Ansari and Asaf Ali. They also made it difficult to oppose and expose the communalism and semi- communalism of leaders like Madan Mohan Malaviya, Lajpat Rai and Maulana Mohammed Ali who often worked within the Congress ranks. 

The strategy of negotiations at the top required generous concessions by the majority to the minority communalism on the question of jobs and seats in the legislatures. But communalism was quite strong among the Hindu middle classes which too suffered from the consequences of colonial underdevelopment. The Congress leadership found it politically difficult to force concessions to Muslim communalism down the throat of Hindu and Sikh communalists. Thus, the failure to conciliate the Muslim communalists helped them gain strength, while any important concessions to them tended to produce a Hindu communal backlash. In any case, even if by a supreme effort in generosity and sagacity a compromise with communal leaders had been arrived at, it was likely to prove temporary as was the case with the Lucknow Pact and to some extent the Nehru Report. Not one communal leader or group or party had enough authority over other communal groups and individuals to sign a lasting agreement. Concessions only whetted the appetite of the communalists. A soon as one group was appeased, a more `extreme' or recalcitrant leader or group emerged and pushed up the communal demands. Consequently, often the more `reasonable' leader or group felt his communal hold over the followers weakening and found it necessary to go back even on the earlier partial or fuller agreement. This is what repeatedly happened during 1928-29 — and Jinnah's was a typical example. The fact was that so long as communal ideology flourished or the socio-political conditions favouring communal politics persisted, it was difficult to appease or conciliate communal leaders permanently or for any length of time. The real answer lay in an all-out opposition to communalism in all arenas — ideological, cultural, social and political. Based on a scientific understanding of its ideology, its social and ideological sources and roots, its social base, and the reasons for its growth in the face of the nationalist work in favour of Hindu-Muslim unity, an intense political-ideological struggle had to be waged against communalism and communal political forces. Moreover, it was necessary to take up the peasants' cause where their class struggle was being distorted into communal channels. All this was not done, despite the deep commitment to secularism of the bulk of the nationalist leadership from Dadabhai Naoroji to Gandhiji and Nehru. 

The need was to direct the debate with the communalists into hard, rational, analytical channels so that the latter were forced to fight on the terrain of reason and science and not of emotion and bias. Gandhiji and the Congress did make Hindu- Muslim unity one of the three basic items of the nationalist political platform. They also, at crucial moments, refused to appease the Hindu communalists. Gandhiji several times staked his life for the secular cause. But Gandhiji and the Congress provided no deeper analysis of the communal phenomenon. 

Despite the intensified activities of communal parties and groups during the 1920s, communalism was not yet very pervasive in Indian society. Communal riots were largely confined to cities and their number, keeping in view the size of the country, was not really large. The Hindu communalists commanded little support among the masses. The social base of the Muslim communalists was also quite narrow. The nationalist Muslims, who were part of the Congress, still represented a major political force. The rising trade union, peasant and youth movements were fully secular. The reaction to the Simon Commission further revealed the weakness of communal forces when both the Muslim League and the Hindu Mahasabha got divided, some in favour of a boycott of the Commission and others for cooperating with it. 

The anti-Simon Commission protest movement and then the Second Civil Disobedience Movement from 1930 to 1934 swept the entire country and once again pushed the communalists as a whole into the background. Led by the Congress, Jamait-ul­ Ulama-i-Hind, Khudai Khidmatgars and other organizations, thousands of Muslims went to jail. The national movement engulfed for the first time two new major areas with a Muslim majority — the North-West Frontier Province and Kashmir. The communal leaders got a chance to come into the limelight during the Round Table Conferences of the early 1930s. At these conferences, the communalists joined hands with the most reactionary sections of the British ruling classes. Both the Muslim and Hindu communalists made efforts to win the support of British authorities to defend their so-called communal interests. In 1932, at a meeting in the House of Commons, the Aga Khan, the poet Mohammed lqbal and the historian Shafaat Ahmad Khan stressed `the inherent impossibility of securing any merger of Hindu and Muslim, political, or indeed social interests' and `the impracticability of ever governing India through anything but a British agency.' Similarly, in 1933, presiding over the Hindu Mahasabha session, Bhai Parmanand made a plea for cooperation between Hindus and the British Government and said: `I feel an impulse in me that Hindus would willingly cooperate with Great Britain if their status and responsible position as the premier community in India is recognized in the political institutions of new India.' 

The communal parties and groups remained quite weak and narrow based till 1937. Most of the Muslim as also Hindu young intellectuals, workers and peasants joined the mainstreams of nationalism and socialism in the early 193 Os. In Bengal, many joined the secular and radical Krishak Praja Party. Moreover, in 1932, in an effort to bolster the sagging Muslim communalism, the British Government announced the Communal Award which accepted virtually all the Muslim communal demands embodied in the Delhi Proposals of 1927 and Jinnah's Fourteen Points of 1929. The communal forces were faced with an entirely new situation; they could not carry on as before. The question was where would they go from here. CHAPTER 33 . 

JINNAH, GOLWALKAR AND EXTREME COMMUNALISM 

Communalism remained at the second, liberal stage till 1937 when it increasingly started assuming a virulent, extremist or fascist form. The liberal communalist argued that India consisted of distinct religion-based communities which had their own separate and special interests which often came into mutual conflict. But he also accepted that the ultimate destiny of Indian politics was the merger of the different communities into a single nation: Thus, the liberal communalist demanded separate communal rights, safeguards, reservations, etc., within the broad concept of one Indian nation-in-the-making. He accepted national unity as the ultimate goal as also the concept of the ultimate common interests of Hindus, Muslims, Sikhs and Christians. Liberal communalism had also a rather narrow social base. Politically, it was based mainly on the upper and middle classes. 

Extreme communalism was based on the politics of hatred, fear psychosis and irrationality. The motifs of domination and suppression, always present in communal propaganda as we have shown earlier, increasingly became the dominant theme of communal propaganda. A campaign of hatred against the followers of other religions was unleashed. The interests of Hindus and Muslims were now declared to be permanently in conflict. The communalists attacked the other `communities' with, in W.C. Smith's words, `fervour, fear, contempt and bitter hatred,' in the extremist or fascist phase of communalism after 1937. Phrases like oppression, suppression, domination, being crushed, even physical extermination and extinction were used. The communalists increasingly operated on the principle: the bigger the lie the better. They poured venom on the National Congress and Gandhiji, and, in particular, they viciously attacked their co-religionists among the nationalists. 

Communalism also now, after 1937, increasingly acquired a popular base, and began to mobilize popular mass opinion. It was now sought to be organized as a mass movement around aggressive, extremist communal politics among the urban lower middle classes. This also required an issue or a slogan which could arouse mass emotion. Because of the reactionary, upper class base of communalism, an appeal to radical social issues could not be made. In other words, communalism could not base itself on a radical socio-economic, or political or ideological programme. Hence, inevitably, an appeal was made to religion and to irrational sentiments of fear and hatred. 

Liberal communalism was transformed into extremist communalism for several reasons. As a consequence of the growth of nationalism and in particular, of the Civil Disobedience Movement of 1930-34, the Congress emerged as the dominant political force in the elections of 1937. Various political parties of landlords and other vested interests suffered a drastic decline. Moreover, as we have seen, the youth as also the workers and peasants were increasingly turning to the Left, and the national movement as a whole was getting increasingly radicalized in its economic and political programme and policies. The zamindars and landlords — the jagirdari elements — finding that open defence of landlords' interests was no longer feasible, now, by and large, switched over to communalism for their class defence. This was not only true in U.P. and Bihar but also in Punjab and Bengal. In Punjab, for example, the big landlords of West Punjab and the Muslim bureaucratic elite had supported the semi- communal, semi-casteist and loyalist Unionist Party. But they increasingly felt that the Unionist Party, being a provincial party, could no longer protect them from Congress radicalism, and so, during the years 1937-45, they gradually shifted their support to the Mus1im League which eagerly promised to protect their interests. Very similar was the case of Muslim zamindars and jotedars in Bengal. Hindu zamindars and landlords and merchants and moneylenders in northern and western India too began to shift towards Hindu communal parties and groups. To attract them, V.D. Savarkar, the Hindu Mahasabha President, began to condemn the `selfish' class tussle between landlords and tenants. Similarly, in Punjab, the Hindu communalists became even more active than before in defending money lending and trading interests. 

Communalism also became, after 1937, the only political recourse of colonial authorities and their policy of divide and rule. This was because by this time, nearly all the other divisions, antagonisms and divisive devices promoted and fostered earlier by the colonial authorities had been overcome by the national movement, and had become politically non-viable from the colonial point of view. The Non-Brahmin challenge in Maharashtra and South India had fizzled out. The Scheduled Castes and other backward castes could no longer be mobilized against the Congress except in stray pockets. The Right and Left wings of the Congress also refused to split. Inter-provincial and inter-lingual rivalries had exhausted themselves much earlier, after the Congress accepted the validity of linguistic states and the cultural diversity of the Indian people. The effort to pit the zamindars and landlords against the national movement had also completely failed. The elections of 1937 showed that nearly all the major social and political groups of colonialism lay shattered. The communal card alone was available for playing against the national movement and the rulers decided to use it to the limit, to stake all on it. They threw all the weight of the colonial state behind Muslim communalism, even though it was headed by a man — M.A. Jinnah — whom they disliked and feared for his sturdy independence and outspoken anti-colonialism. 

The outbreak of World War II, on 1 September, 1939 further strengthened the reliance on the communal card. The Congress withdrew its ministries and demanded that the British make a declaration that India would get complete freedom after the War and transfer of effective Government power immediately. For countering the nationalist demand and dividing Indian opinion, reliance was placed on the Muslim League whose politics and demands were counterposed to the nationalist politics and demands. The League was recognized as the sole spokesperson for Muslims and given the power to veto any political settlement. India could not be given freedom, it was said, so long as Hindus and Muslims did not unite. But such unity was made impossible by the wholesale official backing of Muslim communalism. The Muslim League, in turn, agreed to collaborate with the colonial authorities and serve as their political instrument of its own reasons. The Hindu Mahasabha and other Hindu and Sikh communal organizations also offered to support the colonial Government during the War. But the colonial authorities, while accepting their support, could no longer divide their loyalties; their commitment to Muslim communalism was to remain total during the course of the Wax, and even after. Both the Muslim League and the Hindu Mahasabha had run the election campaign of 1937 on liberal communal lines — they had incorporated much of the nationalist programme and many of the Congress policies, except those relating to agrarian issues, in their election manifestoes. But they had fared poorly in the elections. The Muslim League, for example, won only 109 out of the 482 seats allotted to Muslims under separate electorates, securing only 4.8 per cent of the total Muslim votes. The Hindu Mahasabha fared even worse. 

The communalists now realized that they would gradually wither away if they did not take to militant, mass-based politics. Hitherto, organized mass movements and cadre-based politics had been built by radical, anti-status quo nationalists. The conservatives had shied away from mass movements. In the 1930s, a successful right-wing model of mass politics, which would not frighten away the vested interests, became available in the form of the fascist movement. Both Hindu and Muslim communalists decided to follow this model. Moreover, the Congress had not yet acquired firm roots among all the masses, especially among the Muslim masses; now was the time to take advantage of their political immaturity, before it was too late. Urgency was added to the need to shift to extreme Muslim communalism because the Congress decided to initiate, under Jawaharlal Nehru's guidance, a massive campaign to work among the Muslim masses, known as the Muslim Mass Contact Programme. 

The logic of communalism also inexorably led to extreme communalism. The Congress had gone quite far in the late 1920s in accepting Muslim communal demands. In 1932, the Communal Award and then the Government of India Act of 1935 accepted nearly all the liberal communal demands. Nor did the National Congress oppose these concessions to the communalists. But such concessions would have no cast iron guarantee behind them once the foreign rulers disappeared from the scene and the country came to be ruled democratically. Moreover, what would the communalists do next? Since their demands had been accepted, they had either to dissolve their political organizations, give up communalism and commit political harakiri or discover new demands, new threats to their communities, and inexorably and without necessarily, a conscious design turn towards extreme communalism. Similarly, the Hindu communalists had failed to grow. Further, till 1937, the Congress had permitted both Hindu and Muslim liberal communalists to work within the Congress organization. Under Jawaharlal Nehru's and the Left's pressure the Congress was frontally attacking the communalists. Not only did it not accommodate them in the elections of 1934 and 1937, it moved towards expelling them from the Congress, and finally did so in 1938. The Hindu communalists were facing political extinction. They also had to find a new basis and a new programme for their survival and growth. 

The proposition that communalism has a logic of its own and, if not checked in its early stages, inevitably develops into its `higher' stages is illustrated by the life history of Mohammed All Jinnah. His case shows how communalism is an inclined plane on which a constant slide down becomes inevitable unless counter steps are taken. Once the basic digits of communal ideology are accepted, the ideology takes over a person bit by bit, independent of the subjective desires of the person. This is how a person who started as the `Ambassador of Hindu-Muslim Unity' ended up by demanding Pakistan. 

M.A. Jinnah came back to India after becoming a Barrister in 1906 as a secular, liberal nationalist, a follower of Dadabhai Naoroji. On his return, he immediately joined the Congress and acted as Dadabhai's secretary at the Calcutta session of the Congress in 1906 He was an opponent of the Muslim League then being founded. The Aga Khan, the first president of the League, was to write later that Jinnah was `our toughest opponent in 1906' and that he `came out in bitter hostility toward all that I and my friends had done and were trying to do.. . He said that our principle of separate electorates was dividing the nation against itself.'' From 1906 onwards, Jinnah propagated the theme of national unity in the meetings that he addressed, earning from Sarojini Naidu the title `Ambassador of Hindu- Muslim Unity.' 

The first step towards communalism was taken without any desire of his own and perhaps against his own wishes when he entered the Central Legislative Council from Bombay as a Muslim member under the system of separate electorates. The real slide down began when from a nationalist mire and simple he became a communal nationalist in 1913 when he joined the Muslim League. This, of course, meant that he was still basically a nationalist. He remained in the Congress ad still opposed separate electorates arguing that it would divide India into `two watertight compartments.' But he also started assuming the role of a spokesperson of the Muslim `community' as a whole. These dual roles reached the height of their effectiveness in the Lucknow Congress-League Pact of which he and Tilak were the joint authors. Acting as the spokesperson of Muslim communalism, he got the Congress to accept separate electorates and the system of communal reservations. But he still remained fully committed to nationalism and secular politics. He resigned from the Legislative Council as a protest against the passing of the Rowlatt Bill. He refused the communal assumption that self- government in India would lead to Hindu rule; and argued that the real political issue in India was Home Rule or `transfer of power from bureaucracy to democracy.' 

In 1919-20, the Congress took a turn towards mass politics based on the peaceful breaking of existing laws. Jinnah disagreed and did not find it possible to go along with Gandhi. Along with many other liberals, who thought like him — persons such as Surendranath Banerjea, Bipin Chandra Pal, Tej Bahadur Sapru, C. Sankaran Nair, and many more — Jinnah left the Congress. But he could also see that mere liberal politics had no future. And he was not willing to go into political oblivion. Unlike most of the other liberals, he turned to communal politics. He became a liberal communalist. The logic of communalism had asserted itself and transformed him first from a nationalist into communal nationalist and then into a liberal communalist. 

During the 1920s, Jinnah's nationalism was not fully swallowed by communalism. He revived the down-and-out Muslim League in 1924 and started building it upon and around the demand for safeguarding `the interests and rights of the Muslims.' His politics were now based on the basic communal idea that `Muslims should organize themselves, stand united and should press every reasonable point for the protection of their community.' At the same time, he still pleaded for Hindu-Muslim unity on the basis of a fresh Lucknow Pact so as to fight the British together, and he cooperated with the Swarajists in opposing Government policies and measures in the Central Legislative Assembly. As late as 1925, he told a young Muslim, who said that he was a Muslim first: `My boy, no, you are an Indian first and then a Muslim.' In 1927-28, he supported the boycott of the Simon Commission, though he would not join in the mass demonstrations against it. But by now his entire social base comprised communal- minded persons. He could not give up communalism without losing all political influence. This became apparent in 1928-29 during the discussions on the Nehru Report. Step by step he surrendered to the more reactionary communalists, led by the Aga Khan and M. Shafi, and in the end became the leader of Muslim communalism as a whole, losing in the bargain the support of nationalist leaders like MA. Ansari, T.A.K. Sherwani, Syed Mahmud and his own erstwhile lieutenants like M.C. Chagla. His slide down was symbolized by his becoming the author of the famous 14 demands incorporating the demands of the most reactionary and virulent sections of Muslim communalism. 

Jinnah was further alienated from the main currents of nationalism as the Congress organized the massive mass movement of 1930 and started moving towards a more radical socio-economic programme. Moreover, the Muslim masses especially the younger generation were increasingly shifting to nationalist and left-wing politics and ideologies. Jinnah was faced with a dilemma. He saw little light; and decided to stay mostly in Britain. 

But Jinnah was too much of a man of action and of politics to stay there. He returned to India in 1936 to once again revive the Muslim League. He initially wanted to do so on the basis of liberal communalism. Throughout 1936, he stressed his nationalism and desire for freedom and spoke for Hindu-Muslim cooperation. For example, he said at Lahore in March 1936: `Whatever I have done. let me assure you there has been no change in me, not the slightest, since the day when I joined the 

Indian National Congress. It may be I have been wrong on some occasions. But it has never been done in a partisan spirit. My sole and only object has been the welfare of my country. I assure you that India's interest is and will be sacred to me and nothing will make me budge an inch from that position.' On the one hand, he asked Muslims to organize separately, on the other hand, he asked them to `prove that their patriotism is unsullied and that their love of India and her progress is no less than that of any other community in the country.' 

Jinnah's plan perhaps was to use the Muslim League to win enough seats to force another Lucknow Pact on the Congress. He also assumed that by participating in the 1937 elections the Congress was reverting to pre-Gandhian constitutional politics. Partially because of these assumptions and partially because the bag of communal demands was empty — nearly all the communal demands having been accepted by the Communal Award .Jinnah and the League fought elections on a semi­ nationahst Congress-type of programme, the only `Muslim' demands being protection and promotion of the Urdu language and script, and adoption of measures for the amelioration of the general conditions of Muslims. 

But the poor election results showed that none of Jinnah's assumptions were correct. Jinnah had now to decide what to do: to stick to his semi- nationalist, liberal communal politics which seemed to have exhausted its potentialities or to abandon communal politics. Both would mean going into political wilderness. The third alternative was to take to mass politics which in view of the semi-feudal and semi-loyalist social base of the League and his own socially, economically, and politically conservative views could only be based on the cries of Islam in danger and the danger of a Hindu raj. Jinnah decided in 1937-38 to opt for his last option. And once he took this decision he went all the way towards extreme communalism putting all the force arid brilliance of his personality behind the new politics based on themes of hate and fear. From now on, the entire political campaign among Muslims of this tallest of communal leaders would be geared to appeal to his co-religionists' fear and insecurity and to drive home the theme that the Congress wanted not independence from British imperialism but a Hindu raj in cooperation with the British and domination over Muslims and even their extermination as also the destruction of Islam in India. 

Let us take a few examples. In his presidential address to the League in 1938, Jinnah said: `The High Command of the Congress is determined, absolutely determined to crush all other communities and cultures in this country and establish Hindu raj in this country.' In March 1940, he told the students at Aligarh: `Mr Gandhi's hope is to subjugate and vassalize the Muslims under a Hindu raj.'' Again at Aligarh in March 1941: `Pakistan is not only a practicable goal but the only goal if you wan to save Islam from complete annihilation in this country.'' In his presidential address on April 1941, Jinnah declared that in a united India `the Muslims will be absolutely wiped out of existence.''° Regarding the interim government in 1946, on 18 August, Jinnah referred to `the caste Hindu Fascist Congress,' which wanted to `dominate and rule over Mussalmans and other minor communities of India with the aid of British bayonets.' In 1946, asking Muslims to vote for the League he said: `If we fail to realize our duty today you will be reduced to the status of Sudras and Islam will be vanquished from India.'' 

If a leader of the stature of Jinnah could take up politics and agitation at this low level, it was inevitable that the average communal propagandist would be often even worse. Men like Z.A. Suleri and F.M. Durrani surpassed themselves in Goebbelsian demagogy.' Even Fazl-ul-Huq, holding a responsible position as the Premier of Bengal, told the 1938 session of the League: `In Congress provinces, riots had laid the countryside waste. Muslim life, limb and property have been lost and blood had freely flowed... There the Muslims are leading their lives in constant terror, overawed and oppressed by Hindus.. . There mosques are being defiled and the culprit never found nor is the Muslim worshipper unmolested.'' M.H. Gazdar, a prominent League leader of Sind, told a League meeting in Karachi in March 1941: `The Hindus will have to be eradicated like the Jews in Germany if they did not behave properly.'' Jinnah was however in no position to pull up such people, for his own speeches often skirted the same territory. 

The Muslim communalists now launched a vicious campaign against nationalist Muslims. Maulana Abul Kalam 

Azad and other nationalist Muslims were branded as `show boys' of the Congress, traitors to Islam and mercenary agents of the Hindus. They were submitted, during 1945- 47, to social terror through appeals to religious fanaticism and even to physical attacks. Jinnah himself in his presidential address to the League in April 1943 described Khan Abdul Ghaffar Khan as being `in- charge of the Hinduizing influences and emasculation of the martial Pathans.'' 

Religion was also now brought into the forefront of propaganda. In 1946, Muslims were asked to vote for the League because `a vote for the League and Pakistan was a vote for Islam.' League meetings were often held in the mosques after Friday prayers. Pakistan, it was promised, would be ruled under the Sharia. Muslims were asked to choose between a mosque and a temple. The Quran was widely used as the League's symbol; and the League's fight with the Congress was portrayed as a fight between Islam and Kufr (infidelity). 

Hindu communalism did not lag behind. Its political trajectory was of course different. The two main liberal communal leaders during the 1920s were Lajpat Rai and Madan Mohan Malaviya. Lajpat Rai died in 1928 and Malaviya, finding himself in 1937 in the sort of situation in which Jinnah found himself in the same year, decided to retire from active politics, partly on grounds of health. But Hindu communalism would also not commit suicide; it too advanced to the extremist or the fascist phase. The logic of communalism brought other communal leaders to the fore. The Hindu Mahasabha made a sharp turn in the fascist direction under V.D. Savarkar's leadership. The RSS (Rashtriya Swayamsevak Sangh) had been from the very beginning organized on fascist lines; it now began to branch out beyond Maharashtra. 

Year after year, V.D. Savarkar warned Hindus of the dangers of being dominated by Muslims. He said in 1937 that Muslims `want to brand the forehead of Hindudom and other non-Muslim sections in Hindustan with a stamp of self- humiliation and Muslim domination' and `to reduce the Hindus to the position of helots in their own lands.'' In 1938, he said that `we Hindus are (already) reduced to be veritable helots throughout our land.' 

It was, however, the RSS which became the chief ideologue and propagator of extreme communalism. The head of the RSS, 

M.S. Golwalkar, codified the RSS doctrines in his booklet, We. In 1939, he declared that if the minority demands were accepted, `Hindu National life runs the risk of being shattered.''9 Above all, the RSS attacked Muslims and the Congress leaders. Golwalkar attacked the nationalists for `hugging to our bosom our most inveterate enemies (Muslims) and thus endangering our very existence.'20 Condemning the nationalists for spreading the view by which Hindus `began to class ourselves with our old invaders and foes under the outlandish name — Indian,' he wrote: `We have allowed ourselves to be duped into believing our foes to be our friends ... That is the real danger of the day, our self- forgetfulness, our believing our old and bitter enemies to be our friends.' To Muslims and other religious minorities, Golwalkar gave the following advice: `The non-Hindu peoples in Hindustan must either adopt the Hindu culture and language, must learn o respect and hold in reverence Hindu religion, must entertain no ideas but those of glorification of the Hindu race and culture, i.e., they must not only give up their attitude of intolerance and ungratefulness towards this land and its age long traditions but must also cultivate the positive attitude of love and devotion instead — in one word, they must cease to be foreigners, or may stay in the country, wholly subordinated to the Hindu nation, claiming nothing, deserving no privileges, far less any preferential treatment — not even citizen's rights.' Going further, he wrote: `We Hindus are at war at once with the Muslims on the one hand and British on the other.' He said that Italy and Germany were two countries where `the ancient Race spirit' had `re-risen.' `Even so with us: our Race spirit has once again roused itself,' thus giving Hindus the right of excommunicating Muslims. The RSS launched an even more vicious attack on the Congress leaders during 1946-47. Provocatively accusing the Congress leaders in the true fascist style of asking Hindus to `submit meekly to the vandalism and atrocities of the Muslims' and of telling the Hindu `that he was imbecile, that he had no spirit, no stamina to stand on his own legs and fight for the independence of his motherland and that all this had to be injected into him in the form of Muslim blood', he said in 1947, pointing his finger at Gandhiji: `Those who declared ``No Swaraj without Hindu-Muslim unity'' have thus perpetrated the greatest treason on our society. They have committed the most heinous sin of killing the life-spirit of a great and ancient people.' He accused Gandhiji of having declared: ```There is no Swaraj without Hindu-Muslim unity and the simplest way in which this unity can be achieved is for all the Hindus to become Muslims.'' 

The Hindu communalists also tried to raise the cries of `Hinduism in danger,' `Hindu faith in danger,' and `Hindu culture or sanskriti in danger.' 

The bitter harvest of this campaign of fear and hatred carried on by the Hindu and Muslim communalists since the end of the 19th century, and in particular after 1937, was reaped by the people in the Calcutta killings of August 1946 in which over 5,000 lost their lives within five days, in the butchery of Hindus at Noakhali in Bengal and of Muslims in Bihar, the carnage of the partition riots and the assassination of Gandhiji by a communal fanatic. 

But, perhaps, the heaviest cost was paid by Muslims who remained in or migrated to Pakistan. Once Pakistan was formed, Jinnah hoped to go back to liberal communalism or even secularism. Addressing the people of Pakistan, Jinnah said in his Presidential address to the Constituent Assembly of Pakistan on 11 August 1947: `You may belong to any religion or caste or creed — that has nothing to do with the business of the State... We are starting with this fundamental principle that we are all citizens and equal citizens of one State... Now, I think we should keep that in front of us as our ideal, and you will find that in course of time Hindus would cease to be Hindus and Muslims would cease to be Muslims, not in the religious sense, because that is the personal faith of each individual, but in the political sense as citizens of the State.' But it was all too late. Jinnah had cynically spawned a monster which not only divided India, but would, in time, eat up his own concept of Pakistan and do more harm to Muslims of Pakistan than the most secular of persons could have predicted or even imagined. On the other hand, despite the formation of Pakistan and the bloody communal riots of 1947, nationalist India did succeed in framing a secular constitution and building a basically secular polity, whatever its weaknesses in this respect may be. In other words, ideologies have consequences. 

Two major controversies have arisen in the last thirty years or so around the communal problem. One is the view that the communal problem would have disappeared or been solved if Jinnah had been conciliated during 1937-39 and, in particular, if a coalition government with the Muslim League had been formed in U.P. in 1937. The rebuff to Jinnah's political ambitions, it is said, embittered him and made him turn to separatism. Let us first look at the general argument. It entirely ignores the fact that before he was `rebuffed' Jinnah was already a full- fledged liberal communal 1st. Second, every effort was made by the Congress leaders from 1937 to 1939 to negotiate with Jinnah and to conciliate him. But Jinnah was caught in the logic of communalism. He was left without any negotiable demands which could be rationally put forward and argued. Consequently, and it is very important to remember this historical fact, he refused to tell the Congress leaders what the demands were whose acceptance would satisfy him and lead him to join the Congress in facing imperialism. The impossible condition he laid down to even start negotiations was that the Congress leadership should first renounce its secular character and declare itself a Hindu communal body and accept the Muslim League as the sole representative of the Muslims. The Congress could not have accepted this demand. As Rajendra Prasad put it, for the Congress to accept that it was a Hindu body `would be denying its own past, falsifying its history, and betraying its future' — in fact, it would be betraying the Indian people and their future. If the Congress had accepted Jinnah's demand and `conciliated' him, we might well have been living under a Hindu replica of Pakistan or a Hindu fascist state. So no serious negotiations could even begin. 

Jinnah, too, all the while, was following the logic of his ideology and politics. But this posture could also not be maintained for long. The motive towards Pakistan was then inevitable, for separatism was the only part of the communal ideological programme left unfulfilled. The alternative was to abandon communal politics. And so Jinnah and the Muslim League took the ultimate step in early 1940 and, basing themselves on the theory that Hindus and Muslims were two separate nations which must have separate homelands, put forward the demand for Pakistan. Hindu communalism too had moved in the same direction. Its separatism could not take the form of demanding a part of India as Hindustan — that would be playing into the hands of Muslim communalism. It, therefore, increasingly asserted that Hindus were the only nation living in India and Muslims should either be expelled from India or live in it as second-class citizens. 

Something similar was involved in the U.P. decision of 1937. Jinnah and the League were firmly opposed to mass politics. To have joined hands with them would have meant retreating to constitutional politics in which people had little role to play. Much before the ministerial negotiations occurred or broke down, Jinnah had declared Muslims to be a distinct third party in India, as distinguished from the British and Indian nationalism represented by the Congress. As S. Gopal has put it: `Any coalition with the League implied the Congress accepting a Hindu orientation and renouncing the right to speak for all Indians.'28 It would have also meant the betrayal of nationalist Muslims, who had firmly taken their stand on the terrain of secular nationalism. Furthermore, it would have meant abandonment of the radical agrarian programme adopted at Faizpur in 1936 to which the Congress Ministry was fully committed, for the League was equally committed to the landlords' interests. With their representatives in the Government, no pro-peasant legislation could possibly have been passed. In fact, it was the Congress Socialists and the Communists, quite important in the U.P. Congress at the time, who put pressure on Nehru to reject any coalition with the League and threatened to launch a public campaign on the issue if their demand was rejected. Interestingly, even before negotiations for the formation of a Congress Ministry in U.P. had begun, the Muslim League had raised the cry of `Islam in Danger' in its campaign against Congress candidates in the by-elections to U.P. assembly during May 1937. Jinnah himself had issued appeals to voters in the name of Allah and the Quran. 

In any case, if a leader could turn into a vicious communalist and separatist because his party was not given two seats in a provincial ministry, then how long could he have remained conciliated? To argue in this fashion is, perhaps, to treat history and politics as a joke or as the play of individual whims. The fact is that communalism is basically an ideology which could not have been, and cannot be, appeased; it had to be confronted and opposed, as we have brought out earlier. The failure to do so was the real weakness of the Congress and the national movement. Interestingly, the Communists did try to appease the Muslim League from 1942 to 1946, hoping to wean away its better elements. They not only failed but in the bargain lost some of their best cadres to Muslim communalism. The effort to have a coalition with it turned out to be a one way street from which the Communists had the wisdom to withdraw in 1946. In fact, the negotiations by the Congress leaders as also the Left were based on the false assumption that liberal communalists could be conciliated and then persuaded to fight extreme communalism which was anti-national. After 1937 it was only the nationalist Hindus and Muslims who firmly opposed communalism. Liberal communalists like Malaviya, Shyama Prasad Mukherji and N.C. Chatterjea failed to oppose Savarkar or the RSS. Similarly, the liberal Iqbal or other liberal communal Muslims did not have the courage to oppose the campaign of hatred that Jinnah, Suleri, Fazl-ul-Huq and others unleashed after 1937. At the most, they kept quiet where they did not join it. 

It is also not true that the Congress failure regarding communalism occurred in 1947 when it accepted the partition of the country. Perhaps, there was no other option at the time. Communalism had already advanced too far. There was, it can be argued, no other solution to the communal problem left, unless the national leadership was willing to see the nation plunged in a civil war when the armed forces and the po1ice were under the control of the foreign rulers and were themselves ready to join the civil war. 

The fact is that not all historical situations have an instant solution. Certainly, no such solution existed in 1947. There is never an instant solution to a socio-political problem like communalism. Conditions and forces for a solution have to be prepared over a number of years and even decades. This the Congress and the national movement failed to do. Despite their commitment to secularism, despite Gandhiji's constant emphasis on Hindu-Muslim unity and his willingness to stake his life for its promotion, and despite Nehru's brilliant analysis of the socio­ economic roots of communalism, the Indian nationalists failed to wage a mass ideological-political struggle against all forms of communalism on the basis of patient and scientific exposure of its ideological content, socio-economic roots, and political consequences. In fact, the Congress relied too heavily on negotiations with the communal leaders and failed to evolve a viable and effective long-term strategy to combat communalism at the political, ideological and cultural levels. The Congress and its leadership have to be faulted on this count.
